Automatic Matching of Bullet Lands

### 2015-12-30 imaging-paper.Rnw
reference to supplement

\section{Introduction}
\hh{
Todo list for journal shape-up: there is a lot of good material, we mainly need to focus on what story to tell and make that a solid argument.
\begin{enumerate}
%\item take out a few things: no circular fit, no figures 9 and 10 
%\item do we need the web app?
%\item out: lots of the figures can go into a supplement, together with a discussion. In the paper we then have a summary of what can be seen, so that we can focus on the results. 
%\item add: formula for loess fit and its s.e.
%\item add: comparison of error rates to those of FEs in Hamby study.
%\item change: emphasis in varying cross section. once a stable region is found, how little change is happening? 
\item add: Why do we take the lowest stable cross section?
%\item add: more discussion of the density plots: probability ratio for a match - exchanged probability ratios for ROC curves
\item add: more discussion on additional bullet images. NIST database claims to have 70 bullets -- that would be two sets of the study, but only 35 bullets are currently available. Should we contact Alan Zheng to see whether we can get access to more bullet files?
\item add: should we add a paragraph on the speed of the evaluation? - separate out into pre-processing (needs to be done once) and matching (pairwise comparison of lands). If we only do bullet matching, we could speed this process up by eliminating some land comparisons.
\item add: we might need a proper definition of the joint striation marks (the result of striation-identify)
\item change: we could make the table into a labelled dotplot of importance to save space. We don't need the mean and standard deviations anymore -- the density plots and the ROCs take care of that.
\item change: the focus of the paper is initially quite strongly on the maximal consecutively matching striae (CMS). We need to emphasis the role of CMS in the introduction: max CMS is Biasotti's main criterium for matching.  max CMS is part of current practice (AFTE criteria citation). AFIS \cite{AFIS} is suggesting it for an automated system (was there ever a phase II of that project?). Problem is, that determining CMS is subjective; counting CMS in particular, has been found to be subjective to the examiner: `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}.  For an automatic routine we are faced with the problem of having to determine what a stria is. XXX Should we include one of those microscopic images to bring that point home? We would need somebody qualified to tell us how many CMS there are, though.
\item to consider: AOAS states that accepted papers rarely have more than 20 pages. We are currently at about 23--24 pages.
\item definitions: 
A bullet's {\it signature} is defined as the residual surface measurements after removing the big structure using the loess fit. The {\it cross section} defines the location along the $x$ axis  at which the signature is taken.
\end{enumerate}
}
Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of  convictions  in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\citep{giannelli:2011}.

@@

The smooth curve on the plot represents a segment of a perfect circle with the same radius as the bullet. \hh{While the circle is an obvious first choice for fitting the structure, it does not completely capture the bullet surface after it was fired. A discussion of a circular fit and the remaining residual structure can be found in the supplement.} \hh{XXX Reference to the supplement? XXX}
%It is clear that our next approach must be to model the overall structure in order to focus on the deviations, given that the deviations should correspond to bullet markings.

@@

ggplot() + 
  geom_rect(aes(xmin = xmin, xmax = xmax, fill=factor(type)), ymin = -6, ymax=6.5,  data = lines, alpha=0.2, show.legend = FALSE) +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(lines, match)) +
  ylim(c(-6,6.5)) + theme_bw() +
  geom_line(data=lofX, aes(x=y, y=l30, group=bullet, linetype=bullet)) +
  scale_linetype_discrete("") +
  scale_colour_manual("", values=cols) +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey60", 0.5)) +
    theme(legend.position = c(1,1.1), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab(expression(paste("Signatures (in ",mu,"m)", sep=""))) +
  xlab("y")
@
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between signatures (c).}
\end{figure}
%\afterpage{\clearpage}
Figure~\ref{fig:match} gives an overview of the automated matching routine: 
We first identify a stable region for each bullet and pick the signature at the lowest cross section level in this region, because typically, individual characteristics are best expressed at the lower end of the bullet. 
\hh{(assumption here: lower cross section levels show better expression of individual marks XXX checked XXX crosscuts-and-sd.R)}

@@

\subsection{Impact of cross sections}
The cross section on which a comparison between signatures is based matters -- signatures from cross sections that are further apart, show more pronounced differences between the signatures. 
This poses both a caveat to matching attempts as well as an opportunity \hh{for quality control}: we have to be aware of which cross section level was used in a matching. An assessment of this situation shows that, visually, matches degrade if the signatures upon which the match is based are from cross sections further than 200$\mu m$ apart  (see supplement \hh{XXX section XXX} for a more thorough discussion).
However, we can extract signatures from multiple cross sections of the same bullet land for an initial assessment of its quality. 
This situation is assessed and discussed in more detail below:

@@

\hh{XXX Why do we use the lowest possible cross section? XXX}

@@

aucs <- plyr::ldply(features, function(x) {
  subbstats <- subset(bstats, flagged == FALSE)
  pos.scores <- sample(subbstats[which(subbstats$match),x], 5000, replace=TRUE)
  neg.scores <- sample(subbstats[which(!subbstats$match),x], 5000, replace=TRUE)
  data.frame(x, upper=mean(pos.scores > neg.scores), 
             lower = mean(pos.scores < neg.scores))
})
aucs <- melt(aucs, measure.var=c("lower", "upper"))
names(aucs) <- c("variable", "type", "auc")
rocs <- merge(rocs, aucs, by=c("variable", "type"))
rocs$variable <- reorder(rocs$variable, -rocs$auc, min)

@@

Similarly, the Euclidean distance $D$ between two signatures is strongly correlated to known matches. Out of the 48 pairs of land-to-land signatures with a distance of less than .25, 47 correspond to known matches (the single known mismatch consists of a pair of signatures that both are very flat with peaks and valleys of less than 1.5$\mu m$). %The average distance between known matches in the Hamby study is $\Sexpr{mean(bstats$D[bstats$match])}\mu m$ with a standard deviation of $\Sexpr{sd(bstats$D[bstats$match])}\mu m$; while known mismatches have an average  distance of  $\Sexpr{mean(bstats$D[!bstats$match])}\mu m$ with a standard deviation of $\Sexpr{sd(bstats$D[!bstats$match])}\mu m$.
All of the features in Figure~\ref{fig:densities} show large, if not significant, differences between matches and non-matches. The predictive power of each one of these features is shown in form of ROC curves in Figure~\ref{fig:rocs}. The features are ordered according to the area under the curve (AUC). The feature with the highest individual predictive power is $S$, the sum of two signatures' heights of joint peaks and valleys. The maximal number of CMS is only in the seventh position here. The overall high AUC values indicate, that we can successfully employ machine learning methods to distinguish matches from non-matches.
%Table~\ref{tab:importance} shows an overview in the averages and corresponding standard deviations of the set of feature variables derived in section~\ref{sec:algorithm} from the 3d topological surface measurements. There are quite large differences in these averages, indicating

>>>>

\section{Introduction}
\hh{
Todo list for journal shape-up: there is a lot of good material, we mainly need to focus on what story to tell and make that a solid argument.
\begin{enumerate}
%\item take out a few things: no circular fit, no figures 9 and 10 
%\item do we need the web app?
%\item out: lots of the figures can go into a supplement, together with a discussion. In the paper we then have a summary of what can be seen, so that we can focus on the results. 
%\item add: formula for loess fit and its s.e.
%\item add: comparison of error rates to those of FEs in Hamby study.
%\item change: emphasis in varying cross section. once a stable region is found, how little change is happening? 
%\item add: Why do we take the lowest stable cross section?
%\item add: more discussion of the density plots: probability ratio for a match - exchanged probability ratios for ROC curves
\item add: more discussion on additional bullet images. NIST database claims to have 70 bullets -- that would be two sets of the study, but only 35 bullets are currently available. Should we contact Alan Zheng to see whether we can get access to more bullet files?
\item add: should we add a paragraph on the speed of the evaluation? - separate out into pre-processing (needs to be done once) and matching (pairwise comparison of lands). If we only do bullet matching, we could speed this process up by eliminating some land comparisons.
\item add: we might need a proper definition of the joint striation marks (the result of striation-identify)
\item change: we could make the table into a labelled dotplot of importance to save space. We don't need the mean and standard deviations anymore -- the density plots and the ROCs take care of that.
\item change: the focus of the paper is initially quite strongly on the maximal consecutively matching striae (CMS). We need to emphasis the role of CMS in the introduction: max CMS is Biasotti's main criterium for matching.  max CMS is part of current practice (AFTE criteria citation). AFIS \cite{AFIS} is suggesting it for an automated system (was there ever a phase II of that project?). Problem is, that determining CMS is subjective; counting CMS in particular, has been found to be subjective to the examiner: `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}.  For an automatic routine we are faced with the problem of having to determine what a stria is. XXX Should we include one of those microscopic images to bring that point home? We would need somebody qualified to tell us how many CMS there are, though.
\item to consider: AOAS states that accepted papers rarely have more than 20 pages. We are currently at about 23--24 pages.
\item definitions: 
A bullet's {\it signature} is defined as the residual surface measurements after removing the big structure using the loess fit. The {\it cross section} defines the location along the $x$ axis  at which the signature is taken.
\end{enumerate}
}
Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of  convictions  in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\citep{giannelli:2011}.

@@

The smooth curve on the plot represents a segment of a perfect circle with the same radius as the bullet. \hh{While the circle is an obvious first choice for fitting the structure, it does not completely capture the bullet surface after it was fired. A discussion of a circular fit and the remaining residual structure can be found in supplement~\ref{supp:cylindrical}.} 
%It is clear that our next approach must be to model the overall structure in order to focus on the deviations, given that the deviations should correspond to bullet markings.

@@

ggplot() + 
  geom_rect(aes(xmin = xmin, xmax = xmax, fill=factor(type)), ymin = -6, ymax=6.5,  data = lines, alpha=0.2, show.legend = FALSE) +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(lines, match)) +
  ylim(c(-6,6.5)) + theme_bw() +
  geom_line(data=lofX, aes(x=y, y=l30, group=bullet, linetype=bullet)) +
  scale_linetype_discrete("") +
  scale_colour_manual("", values=cols) +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey60", 0.5)) +
    theme(legend.position = c(1,1.1), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab(expression(paste("Signatures (in ",mu,"m)", sep=""))) +
  xlab("y")
@
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between signatures (c).}
\end{figure}
%\afterpage{\clearpage}
Figure~\ref{fig:match} gives an overview of the automated matching routine: 
We first identify a stable region for each bullet and pick the signature at the lowest cross section level in this region, because typically, individual characteristics are best expressed at the lower end of the bullet \hh{(see supplement~\ref{supp:bulletbottom} for a more detailed discussion)}.

@@

\subsection{Impact of cross sections}
The cross section on which a comparison between signatures is based matters -- signatures from cross sections that are further apart, show more pronounced differences between the signatures. 
This poses both a caveat to matching attempts as well as an opportunity \hh{for quality control}: we have to be aware of which cross section level was used in a matching. An assessment of this situation shows that, visually, matches degrade if the signatures upon which the match is based are from cross sections further than 200$\mu m$ apart  (see supplement~\ref{supp:ccf} for a more thorough discussion).
However, we can extract signatures from multiple cross sections of the same bullet land for an initial assessment of its quality. 
This situation is assessed and discussed in more detail below:

@@

@@

aucs <- plyr::ldply(features, function(x) {
  subbstats <- subset(bstats, flagged == FALSE)
  pos.scores <- sample(subbstats[which(subbstats$match),x], 5000, replace=TRUE)
  neg.scores <- sample(subbstats[which(!subbstats$match),x], 5000, replace=TRUE)
  data.frame(x, upper=mean(pos.scores > neg.scores), 
             lower = mean(pos.scores < neg.scores))
})
aucs <- melt(aucs, measure.var=c("lower", "upper"))
names(aucs) <- c("variable", "type", "auc")
rocs <- merge(rocs, aucs, by=c("variable", "type"))
rocs$variable <- reorder(rocs$variable, -rocs$auc, min)
levels(rocs$variable)[3:4] <- c("#non-matches", "#matches")

@@

Similarly, the Euclidean distance $D$ between two signatures is strongly correlated to known matches. Out of the 48 pairs of land-to-land signatures with a distance of less than .25, 47 correspond to known matches (the single known mismatch consists of a pair of signatures that both are very flat with peaks and valleys of less than 1.5$\mu m$). %The average distance between known matches in the Hamby study is $\Sexpr{mean(bstats$D[bstats$match])}\mu m$ with a standard deviation of $\Sexpr{sd(bstats$D[bstats$match])}\mu m$; while known mismatches have an average  distance of  $\Sexpr{mean(bstats$D[!bstats$match])}\mu m$ with a standard deviation of $\Sexpr{sd(bstats$D[!bstats$match])}\mu m$.
All of the features in Figure~\ref{fig:densities} show large, if not significant, differences between matches and non-matches. The predictive power of each one of these features is shown in form of the Receiving Operating Characteristic (ROC) curves in Figure~\ref{fig:rocs}. The features are ordered according to the area under the curve (AUC). The feature with the highest individual predictive power is $S$, the sum of two signatures' heights of joint peaks and valleys. The maximal number of CMS is only in the seventh position here. The overall high AUC values indicate, that we can successfully employ machine learning methods to distinguish matches from non-matches.
%Table~\ref{tab:importance} shows an overview in the averages and corresponding standard deviations of the set of feature variables derived in section~\ref{sec:algorithm} from the 3d topological surface measurements. There are quite large differences in these averages, indicating
